---
title: "Tara ocean: sparse quadratic log-contrast modeling"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "02-tara-ocean-interactions" # change if you rename file
---

```{r here, message=FALSE, echo = F}
here::i_am(paste0(params$name, ".Rmd"), uuid = "f3ca71d3-1f1c-4612-8aec-d6c86500f6c6")
knitr::opts_chunk$set(dpi = 200, echo = T, warning = F, message = F)
```


```{r packages, echo = F}
library("conflicted")
```

```{r directories, echo = F}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```


# Load Tara Ocean data


This dataset was processed and aggregated in https://github.com/jacobbien/trac-reproducible.

```{r}
path.data <- "data/"
source("R/sparse_log_contrast.R")
source("R/slc_int.R")
source("R/slc_int_plots.R")
source("R/log-ratio-lasso_interactions.R")
source("R/utils.R")
```


```{r}
library(Matrix)
library(tidyverse)
#devtools::install_github("viettr/trac", dependencies = TRUE, ref = "prediction_during_cv")
library(trac)
library(reticulate)
library(ggplot2)
library(gridExtra)
library(ggpattern)
library(ggrepel)
```

```{r}
dat_list <- readRDS(paste0(path.data, "tara_sal_processed_aggregated.RDS"))
```



# 1. SLC and SLC + int.


## Analysis on family level

30 most abundant families


```{r extract-family}
## extract family level
i = "Family"
dat_family <- dat_list[[i]]
X_family <- dat_family$x 
y_sal <- dat_family$y
dim(X_family)
```


```{r}
## keep 30 most abundant families
abund <- colSums(X_family > 0)
X_family <- X_family[, order(abund, decreasing = T)[1:30]]
dim(X_family)
```

```{r rename-family}
rename_X <- stringr::str_split(colnames(X_family), "::")
rename_X <- sapply(rename_X,tail,1)
colnames(X_family) <- rename_X
```

### Make sure that names have no blanks or "-"

```{r rename2-family}
colnames(X_family) <- gsub(" ", "", colnames(X_family))
colnames(X_family) <- gsub("_", "", colnames(X_family))
colnames(X_family) <- gsub("-", "", colnames(X_family))
```


```{r transform-data-family}
pseudo_count <- 1
X_family_psd <- as.matrix(X_family + pseudo_count)
X_family_rel <- X_family_psd/rowSums(X_family_psd)
```



### Fit SLC and SLC + int model

```{r eval = T, results='hide'}
slc_slc_int_family <- slc_slc_int_all_splits(X = as.matrix(X_family_rel),
                                             y = y_sal,
                                             method = "regr", output = "raw")
```



### Compare test and train error


```{r, echo = F}
nsplit = 10
train_mse_slc <- c()
test_mse_slc <- c()

train_mse_slc_int <- c()
test_mse_slc_int <- c()

r_squared_train_slc <- c()
r_squared_test_slc <- c()

r_squared_train_slc_int <- c()
r_squared_test_slc_int <- c()


n <- length(y_sal) - length(slc_slc_int_family$tr[[1]])
p <- ncol(X_family_rel)
ntot <- length(y_sal)
y <- y_sal

for(r in seq(nsplit)){
  yhat_trbest <- slc_slc_int_family$yhat_tr[[r]][, slc_slc_int_family$cvfit[[r]]$cv$i1se]
  yhat_tebest <- slc_slc_int_family$yhat_te[[r]][, slc_slc_int_family$cvfit[[r]]$cv$i1se]
  
  yhat_tr_intbest <- slc_slc_int_family$yhat_tr_int[[r]][, slc_slc_int_family$cvfit[[r]]$cv$i1se]
  yhat_te_intbest <- slc_slc_int_family$yhat_te_int[[r]][, slc_slc_int_family$cvfit[[r]]$cv$i1se]
  
  ytr <- y[slc_slc_int_family$tr[[r]]]
  yte <- y[-slc_slc_int_family$tr[[r]]]
  
  train_mse_slc[r] <- (mean((yhat_trbest - ytr)^2))
  test_mse_slc[r] <- (mean((yhat_tebest - yte)^2))
  
  train_mse_slc_int[r] <- (mean((yhat_tr_intbest - ytr)^2))
  test_mse_slc_int[r] <- (mean((yhat_te_intbest - yte)^2))
  
  
  
  observed_values <- ytr
  predicted_values <- yhat_trbest
  
  # Calculate mean of observed values
  mean_observed <- mean(observed_values)
  
  # Calculate R-squared
  ss_residual <- sum((observed_values - predicted_values)^2)
  ss_total <- sum((observed_values - mean_observed)^2)
  r_squared <- 1 - (ss_residual / ss_total)
  n <- length(ytr)
  r_squared_train_slc[r] <- r_squared
  
  
  observed_values <- yte
  predicted_values <- yhat_tebest
  
  # Calculate mean of observed values
  mean_observed <- mean(observed_values)
  
  # Calculate R-squared
  ss_residual <- sum((observed_values - predicted_values)^2)
  ss_total <- sum((observed_values - mean_observed)^2)
  r_squared <- 1 - (ss_residual / ss_total)
  n <- length(yte)
  r_squared_test_slc[r] <- r_squared
  
  
  observed_values <- ytr
  predicted_values <- yhat_tr_intbest
  
  # Calculate mean of observed values
  mean_observed <- mean(observed_values)
  
  # Calculate R-squared
  ss_residual <- sum((observed_values - predicted_values)^2)
  ss_total <- sum((observed_values - mean_observed)^2)
  r_squared <- 1 - (ss_residual / ss_total)
  n <- length(ytr)
  r_squared_train_slc_int[r] <- r_squared
  
  observed_values <- yte
  predicted_values <- yhat_te_intbest
  
  # Calculate mean of observed values
  mean_observed <- mean(observed_values)
  
  # Calculate R-squared
  ss_residual <- sum((observed_values - predicted_values)^2)
  ss_total <- sum((observed_values - mean_observed)^2)
  r_squared <- 1 - (ss_residual / ss_total)
  n <- length(yte)
  r_squared_test_slc_int[r] <- r_squared
}

# Combine the data into a single data frame
df <- data.frame(
  mse = c(train_mse_slc, test_mse_slc, train_mse_slc_int, test_mse_slc_int),
  group = rep(c("slc train", "slc test", "slc + int train", "slc + int test"), each = length(train_mse_slc))
)

# Add information about the underlying sample size to the group names
df$group <- factor(df$group, labels = c(
  paste0("slc train (n=", n, ")"),
  paste0("slc test (n=", ntot - n, ")"),
  paste0("slc + int train (n=", n, ")"),
  paste0("slc + int test (n=", ntot - n, ")")
))



df <- data.frame(
  r_squared = c(r_squared_train_slc, r_squared_test_slc,
                r_squared_train_slc_int, r_squared_test_slc_int),
  
  group = rep(c("slc train", "slc test", "slc + int train", "slc + int test"), 
              each = nsplit),
  model = rep(c("Main effects model (sparse lc)", "Interaction model (sparse qlc)"), 
              each = nsplit * 2),
  traintest = rep(c("train", "test", "train", "test"), each = nsplit))


# Add information about the underlying sample size to the group names
df$group <- factor(df$group, levels =c("slc train", "slc test", 
                                       "slc + int train", "slc + int test") )
df$model <- factor(df$model, levels =c("Main effects model (sparse lc)", 
                                       "Interaction model (sparse qlc)") )
df$traintest <- factor(df$traintest, levels = c("train", "test"))




# Create the boxplot
plt <- ggplot(df, aes(x = group, y = r_squared, fill = model, pattern = traintest)) +
  geom_boxplot() +
  scale_fill_manual(values = c("lightblue3","steelblue")) +
  geom_boxplot_pattern(position = position_dodge(preserve = "single"), 
                       color = "black", pattern_fill = "white",
                       pattern_angle = 45, pattern_density = 0.1, pattern_spacing = 0.025, 
                       pattern_key_scale_factor = 0.6) +
  scale_pattern_manual(values = c(test = "stripe", train  = "none")) +
  labs(x = "", y = expression("R"^2)) +
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        axis.ticks = element_blank()) +
  guides(
    fill = guide_legend(override.aes = list(pattern = "none"), title = "Model"),
    pattern = guide_legend(override.aes = list(fill = "white"), title = "Train or test data"))

print(plt)
```




```{r}
sum_mat <- rbind(c(mean(train_mse_slc), sd(train_mse_slc)),
                 c(mean(test_mse_slc), sd(test_mse_slc)),
                 c(mean(train_mse_slc_int), sd(train_mse_slc_int)),
                 c(mean(test_mse_slc_int), sd(test_mse_slc_int)))
colnames(sum_mat) <- c("mean MSE", "sd MSE")
sum_mat <- as.data.frame(sum_mat)
sum_mat$model <- c("main-train", "main-test", "int-train", "int-test")
sum_mat$rank <- i


```





```{r}
plt <- list()
for(r in 1:10){
  
  yhat_tebest <- slc_slc_int_family$yhat_te[[r]][, slc_slc_int_family$cvfit[[r]]$cv$i1se]
  yhat_te_intbest <- slc_slc_int_family$yhat_te_int[[r]][, slc_slc_int_family$cvfit[[r]]$cv$i1se]
  
  
  # Create a data frame with the relevant variables
  data <- data.frame(
    yhat_tebest = yhat_tebest,
    yhat_te_intbest = yhat_te_intbest,
    y = y[-slc_slc_int_family$tr[[r]]]  
  )
  
  # Create a scatter plot using ggplot2
  plt[[r]] <- ggplot(data, aes(x = yhat_tebest, y = y)) +
    
    geom_point(data = data, aes(x = yhat_te_intbest, y = y, color = "Interaction model (SLC + int.)"), alpha = 0.8, size = 3) +
    geom_point(aes(color = "Main effects model (SLC)"), alpha = 0.8, size = 3) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "black") +
    labs(title = paste("Train-test split", r),
         x = expression(hat(y)~"(test)"), y = "y (test)") +
    theme_minimal() +
    scale_color_manual(
      values = c("Main effects model (SLC)" = "lightblue3",
                 "Interaction model (SLC + int.)" = "steelblue"),
      name = "Model",
      limits = c("Main effects model (SLC)", "Interaction model (SLC + int.)")
    ) +
    theme(legend.position = "top", legend.justification = "center")
  
  
  
  
} 

plt[[2]]
```



### Plot coefficients

only show coefficients with a nonzero mean estimate.





Main + interaction effects (SLC + int.)


```{r, echo = F}
# Define a vector of names to be replaced and their corresponding replacements
names_to_replace <- c("Surface1", "f2", "f7", "f8", "f13", "f15", "Surface2", 
                      "Surface4", "ChesapeakeDelawareBay", "f26", "f39", "f40", 
                      "f55", "f60", "Deep1", "f96", "f98", "S25593", "JLETNPY6", "ZD0405")
replacements <- c("Surface1 (Order: SAR11 clade)", "f2 (Order: SAR11 clade)", 
                  "f7 (Order: SAR11 clade)", "f8 (Order: SAR11 clade)", 
                  "f13 (Order: SAR11 clade)", "f15 (Order: SAR11 clade)", 
                  "Surface2 (Order: SAR11 clade)", "Surface4 (Order: SAR11 clade)", 
                  "ChesapeakeDelawareBay (Order: SAR11 clade)", "f26 (Order: SAR11 clade)", 
                  "f39 (Order: SAR11 clade)", "f40 (Order: SAR11 clade)", 
                  "f55 (Order: SAR11 clade)", "f60 (Order: SAR11 clade)", 
                  "Deep1 (Order: SAR11 clade)", "f96 (Order: SAR11 clade)", 
                  "f98 (Order: SAR11 clade)", "S25593 (Order: Rickettsiales)", 
                  "JLETNPY6 (Order: Oceanospirillales)", "ZD0405 (Order: Oceanospirillales)")

X_family1 <- X_family
# Iterate over each name to be replaced and perform the replacement
for (i in seq_along(names_to_replace)) {
  colnames(X_family1)[colnames(X_family1) == names_to_replace[i]] <- replacements[i]
}
XX_family1 <- compute.interactions.aitchison(X_family1)
new_name_fam <- c(colnames(X_family1), colnames(XX_family1))
new_name_fam <- gsub("\\(Order:", " (Order:", new_name_fam)
```


```{r echo = F}
mean_ <- apply(slc_slc_int_family$beta_int_est_refit, 1, mean)

# Extract the relevant data
data_to_plot <- t(slc_slc_int_family$beta_int_est_refit)
colnames(data_to_plot) <- new_name_fam
data_to_plot <- data_to_plot[, mean_ != 0]

# Get the order of means
mean_order <- colnames(data_to_plot)[order(apply(data_to_plot, 2, mean))]

# Convert data to long format for ggplot
data_long <- data.frame(value = c(data_to_plot), group = rep(mean_order, each = nrow(data_to_plot)))
# Add a variable indicating whether it's an interaction or linear effect
data_long$effect_type <- ifelse(grepl(":", data_long$group), "Interaction", "Linear")
median_values <- aggregate(value ~ group, data_long, median)
data_long <- merge(data_long, median_values, by = "group", suffixes = c("", "_median"))

# Create ggplot boxplot with different outline colors for interaction and linear effects

coef_int <- ggplot(data_long, aes(x = value, y = factor(group, levels = colnames(data_to_plot)), fill = effect_type)) +
  geom_vline(xintercept = 0, color = "darkgrey") +
  geom_boxplot(position = position_dodge(width = 1), color = "black", fill = "steelblue", width = 0.6) +
  scale_color_manual(values = c("Linear" = "lightblue3", "Interaction" = "steelblue")) +
  labs(y = "", x = expression("Estimated coefficients")) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 14, colour = "black"),
        axis.text.y = element_text(
          angle = 0,
          hjust = 1,
          size = 12,
          colour = "black",
          face = ifelse(colnames(data_to_plot) %in% median_values$group[median_values$value != 0], "bold", "plain")
        ) 
  ) + scale_y_discrete(position = "right")



```

```{r}
median_values$group[median_values$value != 0]
```







## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
