---
title: "04b-ARGs-interactions-binary"
date: "Compiled at `r format(Sys.time(), '%Y-%m-%d %H:%M:%S', tz = 'UTC')` UTC"
output: github_document
params:
  name: "04b-ARGs-interactions-binary" # change if you rename file
---

```{r here, message=FALSE, echo = F}
here::i_am(paste0(params$name, ".Rmd"), uuid = "135a1245-8b68-4cdb-ba43-1b05254a8fdf")
knitr::opts_chunk$set(dpi = 200, echo = T, warning = F, message = F)
```


Here we only use the binarized presence-absence information of the mOTUs.


```{r packages} 
library("conflicted")
library(dplyr)
library(tidyr)
library(textshape)
library(ggplot2)
library(hierNet)
library(gridExtra)
library(glmnet)
library(RColorBrewer)
```

```{r directories, echo = F}
# create or *empty* the target directory, used to write this file's data: 
projthis::proj_create_dir_target(params$name, clean = TRUE)

# function to get path to target directory: path_target("sample.csv")
path_target <- projthis::proj_path_target(params$name)

# function to get path to previous data: path_source("00-import", "sample.csv")
path_source <- projthis::proj_path_source(params$name)
```

### Read data

```{r}
path_data <- "data/"
mOTU_all <- readRDS(paste0(path_data, "mOTU_all.rds"))
meta_all <- readRDS(paste0(path_data, "Metadata_all.rds"))
```


```{r}
## extract genus level and adjust names
mOTU_genus <- mOTU_all$Genus
rownames(mOTU_genus) <- substr(rownames(mOTU_genus), 4, nchar(rownames(mOTU_genus)))

mOTU_genus[1:5, 1:5]
```


Remove samples with missing data in the metadata


```{r}
dim(meta_all)
```

Read ARG data and add it to meta data

```{r}
arg_df <- read.table(paste0(path_data, "hub.microbiome.summary.down.10000000.r"), sep='\t')
arg_df <- arg_df %>% 
  pivot_wider(id_cols = "SampleID", names_from="Feature", values_from = "FeatureValue") %>% 
  select(SampleID, CARD10M) # For now only interested in ARGs

head(arg_df)
```

```{r}
meta_arg <- meta_all %>% 
  tibble::rownames_to_column("SampleID") %>% 
  left_join(arg_df, by="SampleID") %>% 
  column_to_rownames("SampleID")
meta_all <- meta_arg
```


```{r}
meta_all.f = meta_all[complete.cases(meta_all),]
dim(meta_all.f)
```


```{r}
ind_genus = intersect(rownames(meta_all.f), rownames(mOTU_genus))
length(ind_genus)
dim(mOTU_genus)
```

```{r}
## only merged / intersection
meta_all.f.m <- meta_all.f[ind_genus, ]
mOTU_genus.m <- mOTU_genus[ind_genus, ]

dim(meta_all.f.m)
dim(mOTU_genus.m)
```


Remove covariates with only zeros (this step is not really necessary if we only look into ARGs)


```{r}
sum(colSums(meta_all.f.m!= 0) == 0)
meta_all.f.m = meta_all.f.m[, colSums(meta_all.f.m!= 0) > 0]
```

Let's take into account the 30 most abundant genera, remove "unclassified":

```{r}
order_abund <- order(colSums(mOTU_genus.m), decreasing = T)
X <- mOTU_genus.m[, order_abund[2:33]]
## remove duplicates
X <- X[, !(colnames(X) %in% c("_Ruminococcus_", "_Bacteroides_"))]
```

binarize X

```{r}
X[X < 5] <- 0
# X[X < 1] <- 0 # doesn't make a difference on results
X[X > 0] <- 1
table(X)
```

As outcome y we choose the number of ARGs

```{r}
y <- meta_all.f.m$CARD10M
names(y) <- rownames(meta_all.f.m)
all(names(y) == rownames(X))
```


## All pairs-lasso

```{r}
nsplit <- 10
rsq <- function (x, y) cor(x, y) ^ 2

fit.glmnet <- list()
cvfit.glmnet<- list()
yhat_tr.glmnet <- list()
yhat_te.glmnet <- list()
Xall <- cbind(X, hierNet::compute.interactions.c(X, diagonal = F)) 

tr <- list()
ntot <- length(y)
p <- ncol(X)
n <- round(2/3 * ntot)
set.seed(123)
Xm <- X + 0.00001 * matrix(rnorm(length(X)), nrow = nrow(X), ncol = ncol(X))
for(r in seq(nsplit)){
  set.seed(r)
  tr[[r]] <- sample(ntot, n)
  
  fit.glmnet[[r]] <- glmnet(Xall[tr[[r]], ], y[tr[[r]]])
  # set.seed(r)
  cvfit.glmnet[[r]] <- cv.glmnet(Xall[tr[[r]], ], y[tr[[r]]], nfolds = 5)
  yhat_tr.glmnet[[r]] <- predict(cvfit.glmnet[[r]], newx = Xall[tr[[r]], ], s = "lambda.min")
  yhat_te.glmnet[[r]] <- predict(cvfit.glmnet[[r]], newx = Xall[-tr[[r]], ], s = "lambda.min")
}
```

boxplot of coefficients over 10 train-test splits

```{r}
p = ncol(X)
coef_mat_allsplits.glmnet <- matrix(nrow = p * (p - 1) /2 + p, ncol = nsplit)
for(r in seq(nsplit)){
  coef_mat_allsplits.glmnet[, r] <- coef(cvfit.glmnet[[r]], s = "lambda.min")[-1]

}

rownames(coef_mat_allsplits.glmnet) <- colnames(Xall)
coef_mat_allsplits.glmnet <- t(coef_mat_allsplits.glmnet)
# saveRDS(coef_mat_allsplits.glmnet, "coef_APL_binary_ARGs.rds")
```

```{r eval = T, echo = F}
library(ggplot2)
library(reshape2)

# Assuming 'coef_mat_allsplits.glmnet' is already loaded and processed as described
feature_median <- apply(coef_mat_allsplits.glmnet, 2, median)
selected_columns <- coef_mat_allsplits.glmnet[, abs(feature_median) != 0]

s1 <- readRDS("data/selected-features-APL-counts.rds")
s2 <- readRDS("data/selected-features-APL-binary.rds")
s3 <- readRDS("data/selected-features-qlc-compositions.rds")
sel12 <- union(s1, s2)
sel123 <- union(sel12, s3)
non_zero_columns <- coef_mat_allsplits.glmnet[, sel123]

# Get the order from the provided RDS file
order_by_median <- readRDS("data/universal_order_boxplots.rds")

# Prepare data for plotting
data_long <- melt(as.data.frame(non_zero_columns))
data_long$variable <- factor(data_long$variable, levels = rev(order_by_median))

# Assign colors based on the presence of ":"
data_long$color <- ifelse(grepl(":", data_long$variable), "darkred", "lightblue")

# Plotting
plt_coef <- ggplot(data_long, aes(x = value, y = variable, fill = color)) +
  geom_boxplot() +
  labs(x = "Estimated coefficients 10 splits", y = "Features", title = "Boxplot of Estimated Coefficients (10 splits)") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 0, vjust = 0.5, hjust = 1, size = 12, color = "black"),
        axis.text.x = element_text(size = 12, color = "black"),
        legend.position = "none") +
  scale_fill_manual(values = c("lightblue" = "lightblue", "darkred" = "darkred")) +
  xlim(c(-2000, 2000))

# ggsave(paste0(path.temp, "boxplot-coef-MetaCardis-ARGs-qlc-bin.pdf"), plt_coef, height = 10, width = 7)
```


```{r, echo=F, fig.width=15, fig.height=20}
plt_coef
```


```{r}
list_plt_glmnet <- list()
for (r in seq(nsplit)) {
  data <- data.frame(Observed = y[-tr[[r]]], Predicted = as.vector(yhat_te.glmnet[[r]]))
  
  list_plt_glmnet[[r]] <- ggplot(data, aes(y = Observed, x = Predicted)) +
    geom_point(alpha = 0.7, size = 2) +  # Set alpha for transparency
    geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
    labs(title = paste0("Quadratic Lasso (APL): train-test split ", r),
         x = "Predicted number of ARGs (test set)", y = "Observed number of ARGs (test set)") +
    theme_minimal() + xlim(range(c(data$Observed, data$Predicted)))+ 
    theme(axis.text = element_text(color = "black", size = 12), 
          axis.title.y = element_text(color = "black", size = 12),                                            
          axis.title.x = element_text(color = "black", size = 12)
          )
}

# Arrange the ggplots in a single row
# grid.arrange(grobs = list_plt_glmnet[1:2], ncol = 2)
```

```{r}
Rsq <- c()
for(r in seq(nsplit)){
  Rsq[r] <- rsq(y[-tr[[r]]], as.vector(yhat_te.glmnet[[r]]))
}
mean(Rsq)
```


```{r}
list_plt_glmnet <- list()
for (r in seq(nsplit)) {
  data <- data.frame(Observed = y[-tr[[r]]], Predicted = as.vector(yhat_te.glmnet[[r]]))
  
  list_plt_glmnet[[r]] <- ggplot(data, aes(x = Observed, y = Predicted)) +
    geom_point(alpha = 0.5) +  # Set alpha for transparency
    geom_abline(intercept = 0, slope = 1, color = "grey", linetype = "dashed") +
    labs(title = paste0("Quadratic Lasso (APL), presence-absence: train-test split ", r),
         y = "Predicted (test set)", x = "Observed (test set)") +
    theme_minimal() + ylim(range(data$Observed))
}


# Arrange the ggplots in a single row
# grid.arrange(grobs = list_plt_glmnet[1:2], ncol = 2)
```


## Files written

These files have been written to the target directory, ```r paste0("data/", params$name)```:

```{r list-files-target}
projthis::proj_dir_info(path_target())
```
